\subsection{Visual Comparison: Python Code Shape vs. C-Style Braced Code}

Before comparing Python and C at the level of memory management, objects, frames, and runtime execution, it is useful to look at the visible shape of the two languages. Syntax is not only decoration. Syntax is the surface through which a language reveals its assumptions about structure, control, readability, and programmer discipline.

C-style languages usually mark program structure with explicit punctuation. Blocks are enclosed by braces, statements usually end with semicolons, and indentation is a visual convention chosen by the programmer. Python makes a different choice. It removes most of that structural punctuation and turns indentation into part of the grammar. A block is not merely shown by indentation; it is defined by indentation.

Consider a simple conditional in C:

\begin{verbatim}
if (x > 0) {
    printf("positive\n");
} else {
    printf("not positive\n");
}
\end{verbatim}

The same structure in Python is:

\begin{verbatim}
if x > 0:
    print("positive")
else:
    print("not positive")
\end{verbatim}

The visible difference is obvious. C uses parentheses around the condition, braces around the blocks, and semicolons after statements. Python uses a colon to introduce the block and indentation to define the block body. The Python version has less punctuation, but the indentation is no longer optional style. It is syntax.

In C, this is valid even if it is ugly:

\begin{verbatim}
if (x > 0) { printf("positive\n"); } else { printf("not positive\n"); }
\end{verbatim}

The braces still define the structure. The compiler does not care much about the visual layout. In Python, the layout cannot be separated from the structure in the same way:

\begin{verbatim}
if x > 0:
print("positive")
\end{verbatim}

This is not merely badly formatted Python. It is invalid Python, because the body of the \texttt{if} statement must be indented.

The same contrast appears in loops. In C:

\begin{verbatim}
for (int i = 0; i < 10; i++) {
    printf("%d\n", i);
}
\end{verbatim}

In Python:

\begin{verbatim}
for i in range(10):
    print(i)
\end{verbatim}

This example shows two differences at once. The first difference is visual: C uses braces and semicolons, Python uses indentation. The second difference is conceptual: the C loop describes initialization, condition checking, and incrementing directly. The Python loop asks an iterable object, here \texttt{range(10)}, to produce values one by one. The simpler surface syntax is connected to a deeper runtime protocol.

A C programmer may be tempted to read Python as merely ``C without braces.'' That is a mistake. Python does not only replace braces with indentation. It also moves many structural responsibilities into runtime objects and protocols. A Python \texttt{for} loop is not primarily a counter-loop. It is an iteration over an iterable object. A Python function is not only a source block. It becomes a runtime function object. A Python class is not only a static type declaration. It is executable code that creates a class object.

Still, the visual comparison is a good starting point because it shows the first layer of divergence. C makes structure explicit through punctuation. Python makes structure visible through layout.

The difference is especially clear in nested code. C:

\begin{verbatim}
if (user_active) {
    for (int i = 0; i < count; i++) {
        if (items[i].enabled) {
            process(items[i]);
        }
    }
}
\end{verbatim}

Python:

\begin{verbatim}
if user_active:
    for item in items:
        if item.enabled:
            process(item)
\end{verbatim}

The Python version is visually lighter. There are fewer punctuation marks competing for attention. The nesting is shown almost entirely by indentation. This can make code easier to scan, especially for ordinary application logic. The eye follows the left margin and sees the control structure.

But this same design has a cost. In C, braces provide visible block delimiters. A block has an explicit opening and closing marker:

\begin{verbatim}
{
    ...
}
\end{verbatim}

In Python, the block ends when indentation returns to a previous level:

\begin{verbatim}
if condition:
    statement_1
    statement_2
statement_3
\end{verbatim}

Here, \texttt{statement\_3} is outside the \texttt{if} block because it is aligned with the \texttt{if}, not with the indented body. The end of the block is not marked by a closing token. It is marked by dedentation.

This means that Python code shape is more visually honest but also more whitespace-sensitive. If indentation is wrong, the program structure is wrong. In C, bad indentation can mislead the human reader while the braces still control the compiler. In Python, indentation controls both the reader and the parser.

A classic C danger is this:

\begin{verbatim}
if (x > 0)
    printf("positive\n");
    printf("done\n");
\end{verbatim}

Visually, a beginner may think both \texttt{printf} statements belong to the \texttt{if}. They do not. Without braces, only the first statement is controlled by the \texttt{if}. The second statement always runs.

Python avoids that exact mismatch:

\begin{verbatim}
if x > 0:
    print("positive")
    print("done")
\end{verbatim}

Both statements are inside the block because both are indented. If the second statement should always run, it must be dedented:

\begin{verbatim}
if x > 0:
    print("positive")
print("done")
\end{verbatim}

This is Python's strongest argument for indentation-based syntax: the visual structure and grammatical structure are forced to coincide. The program is shaped the way it is read.

However, this does not make Python syntax automatically superior. It only means Python chooses a different enforcement point. C says: use explicit block tokens, and format the code responsibly. Python says: the formatting is the block structure. That is cleaner when code is well written, but stricter and sometimes irritating when code is copied, pasted, generated, or mixed with inconsistent tabs and spaces.

Function definitions show the same contrast. C:

\begin{verbatim}
int add(int a, int b) {
    return a + b;
}
\end{verbatim}

Python:

\begin{verbatim}
def add(a, b):
    return a + b
\end{verbatim}

The C version specifies the return type and parameter types directly in the function signature. The Python version does not. Its parameters are names that will be bound to objects when the function is called. The absence of type declarations is not only a visual simplification. It reflects Python's dynamic binding model. The function does not promise that \texttt{a} and \texttt{b} are machine integers. It promises only that, when the function runs, the operation \texttt{a + b} will be attempted on the objects supplied.

This is why the same Python function may work for several object types:

\begin{verbatim}
add(1, 2)
add("py", "thon")
add([1], [2])
\end{verbatim}

The visible syntax is short because the runtime does the type-dependent work later.

In C, the visible code carries more compile-time information. In Python, the visible code carries less compile-time information and relies more on runtime object behavior. Therefore, the visual simplicity of Python is not free. It is paid for by dynamic dispatch, object metadata, runtime checks, and later error discovery.

A class definition also looks much lighter in Python. C does not have classes in the same sense, but a C-style data structure might look like this:

\begin{verbatim}
struct Point {
    int x;
    int y;
};
\end{verbatim}

Python:

\begin{verbatim}
class Point:
    def __init__(self, x, y):
        self.x = x
        self.y = y
\end{verbatim}

The Python class body is executable code. The method definition creates a function object. The class statement creates a class object. Instance attributes such as \texttt{self.x} and \texttt{self.y} are usually stored as object references in an instance namespace. The C \texttt{struct}, by contrast, defines a fixed memory layout for fields of specified types.

Thus, the visual comparison again points toward a deeper architectural difference. Python looks less rigid because its runtime carries more structure dynamically. C looks more explicit because much of its structure must be known before execution.

A compact comparison is useful:

\begin{verbatim}
C-style code shape:
    braces delimit blocks
    semicolons terminate statements
    declarations often expose storage/type expectations
    indentation is human convention
    compiler relies mainly on punctuation and declarations

Python code shape:
    indentation delimits blocks
    newlines usually terminate statements
    names bind to runtime objects
    indentation is grammar
    interpreter/runtime relies on object metadata and namespaces
\end{verbatim}

The student should therefore avoid two shallow conclusions.

The first shallow conclusion is that Python is just easier because it has less punctuation. That is incomplete. Python has less syntactic punctuation because it moves more responsibility into indentation, object protocols, dynamic lookup, and runtime checks.

The second shallow conclusion is that C is more serious because it has more visible machinery. That is also incomplete. C exposes more low-level structure because it is closer to native compilation and fixed memory layout. Python hides more low-level structure because CPython manages it inside the runtime.

The correct conclusion is that syntax reflects execution philosophy. C source code is shaped around explicit declarations, statement delimiters, braced blocks, and compiled native structure. Python source code is shaped around readable layout, name binding, runtime objects, and interpreter-managed execution.

This visual comparison prepares the next subsection. Once the surface shape is clear, we can compare the architectural reality underneath: Python's high-level managed code model versus C's bare-metal procedural coding model.