\subsection{Architectural Case Study: High-Level Python Managed Code Realities vs. Bare-Metal Procedural C Coding}

The visual difference between Python and C is only the surface. The deeper difference is architectural. C code is written closer to the native process model: typed storage, explicit memory layout, direct control flow, stack-local variables, heap allocation through explicit calls, and compilation toward machine instructions. Python code is written inside a managed runtime model: names bind to objects, objects live on a managed heap, function calls create Python frames, operations dispatch through runtime types, and source code is compiled into bytecode executed by CPython.

A small example is enough to expose the difference. Suppose we want to represent a point and move it.

In C, one natural version is:

\begin{verbatim}
#include <stdio.h>

struct Point {
    int x;
    int y;
};

void move(struct Point *p, int dx, int dy) {
    p->x += dx;
    p->y += dy;
}

int main(void) {
    struct Point p = {2, 3};

    move(&p, 10, 20);

    printf("%d %d\n", p.x, p.y);
    return 0;
}
\end{verbatim}

The structure is explicit. \texttt{struct Point} defines a fixed memory layout containing two integer fields. The local variable \texttt{p} is a concrete object of that structure type, commonly stored in the stack frame of \texttt{main}. The call \texttt{move(\&p, 10, 20)} passes the address of \texttt{p}, so \texttt{move} can modify the original storage. The function receives a pointer, dereferences it, and writes into the two fields.

The programmer can reason in storage terms:

\begin{verbatim}
main stack frame
    p:
        x: int
        y: int

move receives:
    p -> address of that struct
\end{verbatim}

This is a bare-metal procedural style. Data layout is visible. Address passing is explicit. Mutation happens through a pointer. The compiler knows the field offsets, the integer types, and the function signature before execution.

A comparable Python version is:

\begin{verbatim}
class Point:
    def __init__(self, x, y):
        self.x = x
        self.y = y

def move(p, dx, dy):
    p.x += dx
    p.y += dy

p = Point(2, 3)
move(p, 10, 20)

print(p.x, p.y)
\end{verbatim}

The visible code is shorter and less mechanical. There are no explicit pointers, no address operator, no declared integer fields, no header include, no format string, and no manual entry-point function. But this does not mean the machine work disappeared. It moved into the runtime.

When Python executes:

\begin{verbatim}
p = Point(2, 3)
\end{verbatim}

CPython resolves the name \texttt{Point}, calls the class object, allocates a new instance on the managed heap, calls \texttt{\_\_init\_\_}, binds \texttt{self.x} and \texttt{self.y} to integer objects, and then binds the name \texttt{p} to the new instance object.

Conceptually:

\begin{verbatim}
module namespace
    p -> Point instance object
            x -> int object 2
            y -> int object 3
\end{verbatim}

The Python variable \texttt{p} does not contain the point. It refers to a heap object. The attributes \texttt{x} and \texttt{y} are not raw C \texttt{int} fields in the ordinary Python object case. They are attribute bindings pointing to Python integer objects.

When Python executes:

\begin{verbatim}
move(p, 10, 20)
\end{verbatim}

the object referenced by the caller's name \texttt{p} is passed into the function call by binding the parameter name \texttt{p} inside the new function frame to the same object.

During the call:

\begin{verbatim}
module namespace
    p -> Point instance object

move frame
    p  -> same Point instance object
    dx -> int object 10
    dy -> int object 20
\end{verbatim}

This is not C pointer syntax, but it still creates shared access to the same object. The parameter \texttt{p} in \texttt{move} and the global name \texttt{p} both refer to the same heap object. Therefore, modifying attributes through the parameter changes the object observed later through the module-level name.

The line:

\begin{verbatim}
p.x += dx
\end{verbatim}

also hides more work than the C line:

\begin{verbatim}
p->x += dx;
\end{verbatim}

In C, the compiler knows the address of the struct, the offset of field \texttt{x}, and the integer operation required. In Python, CPython must resolve \texttt{p}, look up attribute \texttt{x}, resolve \texttt{dx}, perform Python-level addition according to the runtime types of the objects, and then assign the resulting object reference back to attribute \texttt{x}. Since Python integers are immutable, the operation does not modify the old integer object \texttt{2}. It produces another integer object representing \texttt{12}, and the attribute \texttt{x} is rebound to that result.

So after the first attribute update:

\begin{verbatim}
Point instance object
    x -> int object 12
    y -> int object 3
\end{verbatim}

The instance object was mutated because its attribute binding changed. The integer object was not mutated.

This difference is central. In C:

\begin{verbatim}
p->x += dx;
\end{verbatim}

usually means: find a memory slot inside a struct and modify the integer stored there.

In Python:

\begin{verbatim}
p.x += dx
\end{verbatim}

means: read an attribute reference, compute a new object through Python addition, and store a new reference into the attribute.

The source expressions look similar, but the execution realities are different.

The same contrast appears with collections. In C, an array of integers is usually a contiguous region of raw integer storage:

\begin{verbatim}
int values[3] = {1, 2, 3};
\end{verbatim}

Conceptually:

\begin{verbatim}
values:
    [ int ][ int ][ int ]
\end{verbatim}

In Python:

\begin{verbatim}
values = [1, 2, 3]
\end{verbatim}

the list is a heap object whose element storage contains references to Python objects:

\begin{verbatim}
values -> list object
              slot 0 -> int object 1
              slot 1 -> int object 2
              slot 2 -> int object 3
\end{verbatim}

This is why Python lists can hold mixed objects:

\begin{verbatim}
values = [1, "two", print]
\end{verbatim}

The list does not need one fixed element type because each element slot holds an object reference, and each object carries its own type. The cost is indirection and object overhead. The gain is flexibility.

This case study also explains error timing. In C, many type errors are rejected before execution:

\begin{verbatim}
int x = 10;
x = "hello";
\end{verbatim}

This is invalid because \texttt{x} is an integer storage slot. The compiler can reject the assignment.

In Python:

\begin{verbatim}
x = 10
x = "hello"
\end{verbatim}

there is no type error. The name \texttt{x} is simply rebound from one object to another. But an operation may fail later if the current object does not support the requested behavior:

\begin{verbatim}
x = "hello"
print(x + 5)
\end{verbatim}

The name binding is valid. The addition fails at runtime because a string object and an integer object do not define that operation together.

Thus, C places more responsibility in the compile-time type and storage model. Python places more responsibility in runtime object behavior.

Function calls show the same divergence. In C:

\begin{verbatim}
int add(int a, int b) {
    return a + b;
}
\end{verbatim}

the compiler knows that \texttt{a} and \texttt{b} are integers. The generated code can perform integer addition.

In Python:

\begin{verbatim}
def add(a, b):
    return a + b
\end{verbatim}

the function accepts object references. The operation \texttt{a + b} is resolved when the function runs.

The same function can work for several object types:

\begin{verbatim}
add(1, 2)
add("a", "b")
add([1], [2])
\end{verbatim}

This is useful, but it also means that the runtime must inspect and dispatch operations dynamically. The Python function is more general at the source level because less is fixed before execution.

The architectural contrast can be summarized as follows:

\begin{verbatim}
C:
    variables often correspond to typed storage
    structs define fixed memory layouts
    arrays store raw elements contiguously
    pointers expose addresses explicitly
    many operations compile to direct machine instructions
    memory lifetime may be manual or scope-bound

Python:
    names bind to objects
    objects live on a managed heap
    containers store references to objects
    object types are runtime descriptors
    operations dispatch through type behavior
    memory lifetime follows reference topology
\end{verbatim}

This does not mean Python is merely easier and C is merely harder. They expose different layers of the machine. C exposes storage and native execution more directly. Python exposes program structure, object relationships, and high-level operations more directly. C asks the programmer to manage more low-level detail. Python asks the runtime to manage it, and the programmer must understand the consequences.

The danger for a Python beginner coming from C is to keep using C-shaped explanations. A Python name is not a C variable. A Python list is not a C array. A Python object attribute is not necessarily a fixed struct field. A Python function call is not merely a native jump with typed parameters. A Python class is not merely a compile-time layout declaration.

The danger in the opposite direction is to believe that Python has no memory reality because it hides pointers and allocation. That is also false. Python objects still occupy memory. References still connect objects. Containers still store references. Frames still hold local slots. The runtime still allocates, releases, dispatches, and checks. The machinery is real; it is simply managed by CPython rather than written manually in the source program.

The correct conclusion is that Python is high-level because it relocates low-level responsibilities into the runtime. C is bare-metal in comparison because it leaves more of those responsibilities visible and controllable in the program text. A serious Python programmer does not need to write Python as if it were C, but must understand which parts of C's visible machinery have been moved underneath Python's surface.